Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-20 paper.tex
tratando de generar tex

\date{}
\begin{document}
\maketitle
\begin{abstract}
Education is considered a key tool for social mobility and equality of
opportunities. However, despite the widespread value of education,
disparities in educational outcomes persist over time. To what extent
are such educational disparities justified in society? And, what are the
main factors driving this kind of justification? This research focuses
on the changes in the justification of inequality in access to education
in Chile and their links with the perception of meritocracy. The Chilean
case offers an interesting context for this study given its high
economic inequality, deep neoliberal policies, and commodification of
social services such as education. The central argument of this article
is that the justification of inequalities weakens during periods of
vulnerability and crisis (such as the health and economic crisis
resulting from the COVID-19 pandemic), which could be linked to a
challenge of meritocratic ideals. For testing the research hypotheses we
estimate a series of longitudinal multilevel models with data from a
Chilean longitudinal panel survey (2016 -- 2022, 6 waves, N = 2,927).
Whereas we find support for the association between the perception of
meritocracy and justification of inequality, the analyses that involved
changes over time showed an increase in the justification of inequality
in education. The discussion of these results delves into the social
consequences of justifying inequality in a sensitive area as education,
as well as the persistence of meritocratic ideals despite challenging
events.

\newline
\newline

\textbf{Keywords}: meritocracy, social inequality, inequality
justification, COVID-19
\end{abstract}

This document was last modified at 2024-12-15 23:10:25

and it was last rendered at 2024-12-15 23:10:25

@@

Despite the widespread recognition of education as a key factor for
social mobility and equality of opportunities, disparities in this
domain are far from being eliminated. As the last PISA Report points
out, ``Socio-economically disadvantaged students are seven times more
likely than advantaged students to score below Level 2 in mathematics on
average across OECD countries'' (OECD 2023). Furthermore, empirical
evidence from social science research has revealed an exacerbation of
educational inequalities across various domains, such as family
socioeconomic status, gender, ethnicity, and migratory background,
particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
Nevertheless, and despite the consistent empirical evidence on
educational inequalities and their consequences, research on
distributive justice indicates remarkable individual differences in the
justification of unequal access to education (Lee and Stacey 2023). In
the present paper, we argue that the perception of meritocracy in
society might play a role in the justification of economic inequalities
in this particular domain.

Meritocracy is defined as a distributive system where individual factors
such as effort and talent are conceived as the main determinants of
individual performance, usually sidelining the role of opportunities and
social background. In recent years, there has been a growing interest in
studying meritocratic beliefs as well as their variations according to
individual and contextual variables (Mijs 2021; Mijs et al. 2022). This
line of research has highlighted how meritocratic beliefs are associated
with justifying unequal access to resources and rewards in society, by
legitimizing social hierarchies in which those who already have
resources and advantages are more likely to succeed (Sandel 2020;
McNamee and Miller 2004; Hadjar 2008). Within this framework, this study
focuses on the changes in the justification of inequalities in access to
education in Chile, and to what extent this phenomenon is associated
with the perception of meritocracy and its variations during the COVID
pandemic.

The Chilean case is particularly relevant when it comes to the study of
inequalities in education. The country is characterized by a high level
of economic inequality associated with the privatization and
commodification of several social policy domains, among them the
educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
2016). The neoliberal reforms implemented during and since the military
dictatorship (1973-1989) have deepened educational policies based on
private subsidies, achievement incentives, selection, and segregation
(Joiko 2019). Previous research in Chile has emphasized the concept of
an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
2016), characterized by selection, competition, and the provision of
goods and services associated with payment capacities. Many studies have
addressed this problem from economic, political, and sociological
perspectives, generating relevant evidence on their academic and social
consequences (Ramos 2022). Likewise, massive social mobilizations in the
country (particularly in 2006, 2011, and 2018) have had educational
inequalities as one of the main targets, putting pressure on governments
towards a series of reforms that have advanced free higher education
reforms and have diminished the role of private actors in the
educational system.

This research endeavors to dissect the nuanced alterations in the
justifications for educational disparities within the Chilean framework.
Central to our investigation is the premise that the rationale
underpinning these inequalities experiences a significant recalibration
during periods of vulnerability and crises, specifically under the
unprecedented strains of the global health and economic cataclysm
triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
Breznau 2021). A pertinent aspect of this discourse is articulated in
the works of Mijs (2016, 2018), showing the interplay between
meritocratic beliefs and the legitimization of disparities, as well as
the potential erosion of meritocratic principles amidst crises. Within
this framework, we point out that periods of vulnerability (such as the
health and economic crisis resulting from the COVID-19 pandemic) could
weaken the justification of social inequalities, a change that would be
at least in part linked to a challenge of the meritocratic ideals.
Furthermore, we suggest that the association between meritocratic
perception and inequality justification would be weaker among
lower-status groups, as they are the most affected in critical times.

\section{Inequality, beliefs and
justification}\label{inequality-beliefs-and-justification}

The justification of economic inequalities is a controversial field of
study. As equality is usually considered a synonym of justice in common
sense, the idea that someone could be willing to justify inequalities
seems (in principle) far from what is considerably reasonable.
Nevertheless, and as pointed out by Barry, the whole normative
discussion in social justice can be conceived as ``the defensibility of
unequal relations between people'' (1989, 3), whereby empirical research
has consistently shown strong variations between individuals and
societies regarding inequality justification (Trump 2018; Kelley and
Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
been able to establish a clear difference between inequality and
justice, which opens the question of to what extent inequality can be
tolerated or even justified.

The field of studies of subjective economic inequality has been
extensively developed in the social sciences, particularly in the area
of social justice research (Resh and Sabbagh 2016). From a sociological
point of view, the works of Kluegel and Smith (1981) can be considered
pioneers in framing a research area comprising perceptions, beliefs, and
justifications in topics such as inequality of opportunities, poverty,
and distributive justice. Theoretically, the authors understand the
concept of ``beliefs'' in a ``broad sense to refer to the information
(veridical or non-veridical) about a phenomenon that an individual uses
as a basis both for inferring other information and for action.''
(Kluegel and Smith 1981, 30). However, the literature has strained the
concept of ``beliefs'' by more specific distinctions around the
individual and structural dimensions of economic inequality. In this
regard, Wegener (1987) has suggested that perceptions of social status
and reward hierarchy are relative to the observer's status in such
hierarchies (Wegener 1987, 2). This approach established that status
positions influence how people perceive things \emph{are} and, in turn,
these perceptions anchor how people evaluate this inequality as just or
unjust (Jasso 1978). Additionally, Janmaat (2013) has suggested more
detailed conceptual distinctions about how people perceive economic
inequality \emph{is}, how they believe or prefer it \emph{should be},
and the process of evaluating such differences as fair.

Theoretically, Liebig and Sauer (2016) establish three fundamental
distinctions in justice attitudes in the field of sociological justice
research. First, \emph{order-related attitudes} refer to individual
motives that explain preferences for resource allocation based on the
principles of equality, need, or entitlement. Second, \emph{procedural
justice attitudes} account for the degree of acceptance of unfavorable
outcomes under conditions in which the procedure under which those
decisions were made is considered fair (e.g., opportunities). Finally,
they can be \emph{outcome-related attitudes} which refer primarily to
the economic resources that a person currently obtains and the set of
collective commitments associated with possessing those resources (e.g.,
transfers or taxes). Within this framework, we conceptualize the
justification of inequality in the first area of order-related
attitudes, as it expresses preferences for resource allocation that can
be associated with concepts that reflect different distributive
principles (Kluegel, Mason, and Wegener 1995; Son Hing et al. 2011)

An important finding in the area of order-related attitudes is that
beliefs about inequality are stratified by social status (Janmaat 2013;
Kluegel and Smith 1986). The evidence suggests that advantaged positions
in terms of socioeconomic status, such as income, education, or
occupational prestige, are associated with preferences for less
egalitarian class imageries (Evans, Kelley, and Kolosi 1992; Evans and
Kelley 2017) and greater justification of salary gaps (Hadler 2005;
Kelley and Evans 2021). Nevertheless, and in contrast to this
perspective, it has been argued that human beings are intrinsically
motivated to justify the system they inhabit, even when the information
available to them points in the opposite direction (Jost and Hunyady
2003; van der Toorn and Jost 2014). Individuals would be generally
disposed to justify the economic status quo regardless of self- and
group economic interest because their support partially depends on their
beliefs in system-justifying ideologies (Trump 2020). As a result, what
has been suggested is that low-status might perceive the status quo as
just as they are more strongly attached to dominant ideologies, while
high-status individuals are more prone to justify the system to a
greater extent because it favors their interests and dominant position
(Jost 2020).

Another factor that has been linked to inequality beliefs is the
perception of economic inequality (Kelley and Evans 1993; Osberg and
Smeeding 2006; Castillo, García-Castro, and Venegas 2022). Studies in
this matter have argued that perceived economic inequality functions as
an anchor for the justifications of inequality because if these
inequalities are indeed seen as fair in their generative processes, then
they will be considered as just (Wegener 1987, 1990). Thus, both the
role of social status and economic inequality perceptions have found
support in correlational (Castillo 2012; Schröder 2017), as well as in
experimental studies that have manipulated inequality perception (Iturra
et al. 2023; Trump 2018). In addition, normative or value-driven
mechanisms can also explain to what extent people justify inequality.
For instance, studies suggest that justice ideologies such as
individualism or egalitarianism can mediate the influence of individual
status on income inequality justification (Castillo 2011). Likewise,
system justification ideologies can strengthen the influence of
socioeconomic status on economic inequality preferences (García-Sánchez
and De Carvalho 2022). Altogether, these studies have shown that not
only socio-structural factors influence what people believe regarding
the distribution of economic resources, but their normative stances and
perceptions can also determine preferences for economic inequality in
diverse domains.

\section{The justification of educational
inequality}\label{the-justification-of-educational-inequality}

@@

The study of inequality justification in the sphere of education is a
rather specific and unattended field. Most studies on educational
inequality relate performance indicators (such as grades and
standardized test results) to stratification-related variables, whereby
few refer specifically to if and how individuals tolerate and even
justify educational inequalities. This is probably because in principle
it seems counter-intuitive to think that someone would be willing to
justify differences in such a sensitive area. This paradox is phrased by
Conforti (1992) as ``while equality is embraced as a value, inequality
is accepted as a reality'' (p.230).

In the area of empirical studies, the operationalization of educational
inequality justification in terms of questionnaire items is diverse, and
there is not a single instrument or scale dedicated specifically to this
concept. Rather, some attitudinal surveys usually include
education-related items that are posteriorly used as dependent or
independent variables. Some examples are the International Social Survey
Programme (ISSP) (\emph{Is it just or unjust -- right or wrong -- that
people with higher incomes can buy better education for their children
than people with lower incomes?}), and the European Social Survey (ESS)
2018 (\emph{Compared to other people in {[}country{]}, I have had a fair
chance of achieving the level of education I was seeking}). Although
both questions have been used to measure the justification of
educational inequality, they refer to different aspects. The ESS can be
classified as reflexive, as the respondent is involved as a referent for
the question, whereas the ISSP question is non-reflexive. Furthermore,
ISSP refers to educational access, while the ESS refers to educational
outputs.

Studies have addressed the justification of educational inequality from
different conceptual perspectives and methodological strategies. Some
perspectives emphasize the justification of differences between specific
groups, finding for instance that Americans are more concerned
about---and more supportive of proposals to close---wealth-based
achievement gaps than Black-White or Hispanic-White gaps (Valant and
Newark 2016). Other studies focus on the association between status
differences and inequality justification in education, based on the
rational premise that those who are worst-off would tend to show lower
levels of justification (Stoilova and Ilieva-Trichkova 2023). However,
most of the research in this area goes beyond status differences,
focusing on how subjective factors such as perceptions and beliefs are
related to different levels of justification.

The perception of economic inequality is one of the factors associated
with the justification of educational inequality. As pointed out by
Janmaat (2013), justifications are usually confused under the labels of
``beliefs'' and ``attitudes,'' and their distinction from perceptions is
relevant as it helps state that justifying a certain level of inequality
requires taking into account which level of (perceived) inequality is
first and foremost being recognized (Son Hing et al. 2011; Castillo et
al. 2023). For instance, a recent study by Day and Norton (2023) on the
perception and justification of inequality in university endowments in
the U.S. points out that individuals tend to underestimate the magnitude
of existing inequality while simultaneously desiring greater equality.
Furthermore, through an experiment, they demonstrate that information
about inequality in university endowments (i.e., manipulating inequality
perception) increases the perception of injustice in education. Results
of this study are in line with the consistent evidence on biases related
to the underestimation of economic inequality (Castillo, García-Castro,
and Venegas 2022; Gimpelson and Treisman 2018)

Besides focusing on the relationship between perception and
justification of educational inequality, another group of studies has
addressed how the justification of inequality is associated with the
support of different distributive principles. In an experiment by
Igliozzi, Granot, and Ottati (2024), they establish how the support for
different justice principles (equity, equality, and need) is related to
support for redistributive policies in domains such as education. They
find that a larger system justification (Jost and Hunyady 2003) is
related to higher support for the principles of equity and equality
(over need) in distributing educational outcomes, where equity was
measured in terms of access to education related to performance
(`Students who place within the highest 50\% of testing scores from the
previous year can attend the magnet school'). In the same line, Lee and
Stacey (2023) show that performance assessments (in what they call
``neoliberal orientations'') predicted people's fair perceptions of
educational inequality.

\emph{Justificacion, inequality and meritocracy in the educational
domain}

The relationship between performance and outputs brings us to the
concepts of merit and meritocracy, key in educational justice research.
The concept of meritocracy, coined by the sociologist Michael Young in
his satirical work ``The Rise of the Meritocracy'' (1962), has a rich
history and multifaceted meaning. Young's term was initially used to
critique a hypothetical future society where social and economic
positions were entirely determined by individuals' merits, talents, and
efforts, ostensibly eliminating the influence of inherited privilege.
However, over time, the term has evolved to describe a social system
where advancement and success are perceived as directly linked to one's
abilities and achievements, sidelining inequalities of opportunities
(Bell 2020). This would contribute to the justification of inequality in
the educational domain (Sandel 2020), as individual ability is often
inherited, and educational attainment is influenced by family background
(Goldin 2000).

Empirical research examining the relationship between meritocratic
perceptions and the justification of educational inequality provides
valuable insights into the dynamics shaping individuals' attitudes and
beliefs. Some studies have explored the extent to which people believe
that educational systems are meritocratic and how these perceptions
influence their acceptance or rejection of educational disparities, as
well as status differences (McCoy and Major 2007). This belief is
particularly influential in educational institutions, where it not only
encourages academic success but also legitimizes social and income
inequality (Batruch et al. 2023). Most of the evidence so far suggests
that those who believe in a direct link between effort, talent, and
educational success are more inclined to rationalize existing
disparities in educational outcomes. However, empirical findings also
reveal nuances in this relationship. For instance, studies show that
when individuals perceive structural barriers, such as economic
disparities or discrimination, as hindering the meritocratic ideal, they
may be less likely to justify educational inequalities (Day and Norton
2023). This suggests that while meritocratic beliefs play a significant
role, individuals' awareness of broader societal issues can mediate
their willingness to rationalize educational disparities.

\emph{This study}

The present research uses longitudinal survey data from 2016 to 2022 to
asses changes in inequality justification. As there are measures before
and after the COVID pandemic, firstly we explore to what extent it is
possible to identify variations in inequality justification in this time
frame. Although we are aware that it is not possible to make a causal
claim given that this period is characterized by many other
circumstances besides COVID, we argue that the increase of risk
perceptions in critical situations (such as the pandemic) impacts public
concerns about equality and redistribution (Breznau 2021). Although the
pandemic might have triggered such concerns mostly regarding health, we
believe that other policy areas, such as education, could have been
impacted in this period. Consistently, our first hypothesis is:

@@

Our second argument is based on the evidence that relates rational
interests with justifications and preferences (Son Hing et al. 2019),
expecting that:

\(H_2\): Those with lower status characteristics show lower
justification of educational inequalities

Thirdly, we address inequality perception. As referred to above, the
justification of inequality requires, first and foremost, that such
inequality is recognized. As a larger inequality in this domain
threatens a basic equality ideal, the corresponding hypothesis in this
case is:

\(H_3\): A larger inequality perception is negatively associated with
the justification of educational inequalities.

Our fourth hypothesis refers to the central concept of meritocracy. As
meritocracy justifies inequalities based on performance, we expect that:

@@

Finally, and in line with hypotheses 1 and 4, we explore interactions
between meritocracy and inequality justification over time. As COVID-19
pandemic has affected the structural opportunities to attend schools,
universities, or workplaces, the ideal of meritocracy might have been
challenged. As a consequence, the justification of inequality could show
a decrease over time:

@@

@@

The main data source is the Chilean Longitudinal Social Survey
2016--2022. The ELSOC has been designed as a yearly panel study to
evaluate how individuals think, feel, and behave regarding a set of
social issues concerning conflict and social cohesion in Chile. The
sampling design is probabilistic, stratified, clustered, and multistage.
It provides adequate coverage of the country's largest cities
(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
cities. Consequently, the ELSOC comprises a total of 2,927 participants
aged between 18 and 75 years in wave 1. In addition, the sample
represents 77\% of the total national population and 93\% of the urban
areas (ELSOC 2022).

The survey has been conducted yearly since 2016, except for the year
2020, when it was suspended due to the pandemic. The waves 2016, 2017,
2018, 2019 and 2022 were administered using computer-assisted personal
interviewing (CAPI). However, a reduced version was conducted using
computer-assisted telephone interviews (CATI) in 2021. In addition, wave
3 included a refreshment sample to counter survey attrition. The same
sampling strategy as that of wave 1 was implemented for selecting new
cases. As a result, the total sample of wave 3 included 3,748 cases, of
which 2,229 were part of the original sample and 1,519 were from the
refreshment sample. The data from the refreshment sample is not included
in the analytical sample because we wanted to analyze longer response
trends. Regarding the original sample, the response rate was 62.4\% in
wave 1, achieving N = 2,927 participants. In broader terms, the
accumulated attrition between wave 1 and wave 6 is 41\%, achieving a
final sample of N = 1728. The analysis used longitudinal weights to
avoid bias due to the sampling design, allowing us to control for biases
arising from systematic patterns of non-participation in the survey
after the first wave. For a more detailed analysis of responses,
attrition, and the construction of longitudinal weights, visit
https://coes.cl/encuesta-panel/. The dataset is publicly available at:
https://dataverse.harvard.edu/dataverse/elsoc.

@@

For testing the hypothesis regarding status we included educational
level and household income quintiles, as well as subjective social
status. Finally, political identification on a left-right scale, age,
and gender were included as control variables.
Table~\ref{tbl-descriptives} shows the whole set of independent
variables, the corresponding items, response categories, and
frequencies.

>>>>

\date{}
\begin{document}
\maketitle
\begin{abstract}
Education is considered a key tool for social mobility and equality of
opportunities. However, despite the widespread value of education,
disparities in educational outcomes persist over time. This paper is
guided by the following research questions: To what extent are such
educational disparities justified in society? Is the justification of
educational inequality affected in periods of personal and social
vulnerability, as in the COVID pandemic? And, what are the main factors
driving this kind of justification? The Chilean case offers an
interesting context for this study given its high economic inequality,
deep neoliberal policies, and commodification of social services such as
education. This economic and cultural environment has promoted
meritocratic ideals, whereby individual talent and effort are considered
key to get ahead in life, disregarding opportunities linked to social
origin. The central argument of this article is that the justification
of inequalities weakens during periods of vulnerability and crisis (such
as the health and economic crisis resulting from the COVID-19 pandemic).
Furthermore, we argue that such changes could be linked to a challenge
of meritocratic ideals. For testing the research hypotheses we estimate
a series of longitudinal multilevel models with data from a Chilean
longitudinal panel survey (2016 -- 2023, 6 waves, N = 2,927). Whereas we
find support for the association between the perception of meritocracy
and justification of inequality, the analyses that involved changes over
time showed an increase in the justification of inequality in education.
The discussion of these results delves into the social consequences of
justifying inequality in a sensitive area as education, as well as the
persistence of meritocratic ideals despite challenging events.\newline
\textbf{Keywords}: meritocracy, social inequality, inequality
justification, COVID-19
\end{abstract}

This document was last modified at 2025-01-19 22:27:33

and it was last rendered at 2025-01-19 22:27:33

@@

Despite the widespread recognition of education as a key factor for
social mobility and equality of opportunities, disparities in this
domain are far from being eliminated. As the last PISA report points
out, ``Socio-economically disadvantaged students are seven times more
likely than advantaged students to score below Level 2 in mathematics on
average across OECD countries'' (OECD 2023). Furthermore, empirical
evidence from social science research has revealed an exacerbation of
educational inequalities across various domains, such as family
socioeconomic status, gender, ethnicity, and migratory background,
particularly after COVID-19 (Darmody, Smyth, and Russell 2021).
Nevertheless, and despite the consistent empirical evidence on
educational inequalities and their consequences, research on
distributive justice indicates remarkable individual differences in the
justification of unequal access to education (Lee and Stacey 2023). In
the present paper, we argue that the perception of meritocracy in
society might play a role in the justification of economic inequalities
in this particular domain. We examine this association in the context of
COVID 19 in Chile, a period characterized by several challenges
regarding the political and economic system that could have affected the
way in which the role of individual performance and social outcomes are
deemed.

The Chilean case is particularly relevant when it comes to the study of
inequalities in education. The country is characterized by a high level
of economic inequality associated with the privatization and
commodification of several social policy domains, among them the
educational system (Bellei 2013; Corvalán, Carrasco, and García-Huidobro
2016). The neoliberal reforms implemented during and since the military
dictatorship (1973-1989) have deepened educational policies based on
private subsidies, achievement incentives, selection, and segregation
(Joiko 2019). Previous research in Chile has emphasized the concept of
an ``educational market'' (Corvalán, Carrasco, and García-Huidobro
2016), characterized by selection, competition, and the provision of
goods and services associated with payment capacities. Many studies have
addressed this problem from economic, political, and sociological
perspectives, generating relevant evidence on their academic and social
consequences (Ramos 2022). Likewise, massive social mobilizations in the
country (particularly in 2006, 2011, and 2018) have had educational
inequalities as one of the main targets, putting pressure on governments
towards a series of reforms that have advanced free higher education
reforms and have diminished the role of private actors in the
educational system (Donoso 2016; Reyes-Housholder and Roque 2019).

This research endeavors to dissect the nuanced alterations in the
justifications for educational disparities within the Chilean framework.
Central to our investigation is the premise that the rationale
underpinning these inequalities experiences a significant recalibration
during periods of vulnerability and crises, specifically under the
unprecedented strains of the global health and economic cataclysm
triggered by the COVID-19 pandemic (Hankivsky, Morrow, and Varcoe 2022;
Breznau 2021). A pertinent aspect of this discourse is articulated in
the works of Mijs (2016, 2018), showing the interplay between
meritocratic beliefs and the legitimization of disparities, as well as
the potential erosion of meritocratic principles amidst crises. This
line of research has highlighted how meritocratic beliefs are associated
with justifying unequal access to resources and rewards in society, by
legitimizing social hierarchies in which those who already have
resources and advantages are more likely to succeed (Sandel 2020;
McNamee and Miller 2004; Hadjar 2008). Within this framework, we point
out that periods of vulnerability (such as the health and economic
crisis resulting from the COVID-19 pandemic) could weaken the
justification of social inequalities, a change that would be at least in
part linked to a challenge of the meritocratic ideals.

\section{The justification of educational
inequalities}\label{the-justification-of-educational-inequalities}

The justification of economic inequalities is a controversial field of
study. As equality is usually considered a synonym of justice in common
sense, the idea that someone could be willing to justify inequalities
seems (in principle) far from what is considerably reasonable.
Nevertheless, and as pointed out by Barry, the whole normative
discussion in social justice can be conceived as ``the defensibility of
unequal relations between people'' (1989, 3), whereby empirical research
has consistently shown strong variations between individuals and
societies regarding inequality justification (Trump 2018; Kelley and
Evans 2021; Mijs 2021). Both theoretical and empirical approaches have
been able to establish a clear distinction between inequality and
justice, opening the question of to what extent inequality can be
tolerated or even justified.

@@

In the area of survey studies, the operationalization of educational
inequality justification in terms of questionnaire items has been
diverse, and there is not a single instrument or scale dedicated
specifically to this concept. Rather, some attitudinal surveys usually
include education-related items that are posteriorly used as dependent
or independent variables. Some examples are the International Social
Survey Programme (ISSP) (\emph{Is it just or unjust -- right or wrong --
that people with higher incomes can buy better education for their
children than people with lower incomes?}), and the European Social
Survey (ESS) 2018 (\emph{Compared to other people in {[}country{]}, I
have had a fair chance of achieving the level of education I was
seeking}). Although both questions have been used to measure the
justification of educational inequality, they refer to different
aspects: the ESS can be classified as reflexive, as the respondent is
involved as a referent for the question, whereas the ISSP question is
non-reflexive. Furthermore, ISSP refers to educational access, while the
ESS refers to educational outputs. Although such items covering
inequality justification are present in emblematic comparative studies,
still there are few studies focusing specifically in the justification
of educational inequality.

\emph{The COVID pandemic and public concerns about inequality}

The time frame for this study covers between 2016 and 2023, by using
longitudinal panel survey data provided by the ELSOC database (Estudio
Longitudinal Social de Chile). Although this period comprises the COVID
pandemic, the changes in the variables analyzed are not possible to be
attributed only to this event in terms of a health crisis, as several
economic, social and cultural processes overlapped worldwide. However,
and as several studies have shown, this period appeared to trigger
attitudinal changes that are relevant to explore, both in themselves and
as antecedent to future studies. In this sense, in this paper we argue
that the increase of risk perceptions in critical situations (such as
the pandemic) impacts public concerns about equality and redistribution
(Breznau 2021).

In Chile, the COVID-19 pandemic posed a significant threat to
economically vulnerable groups and those in precarious labor market
positions (Ministerio de Desarrollo Social y Familia 2023; Instituto
Nacional de Estadísticas 2022). The poverty rate, which was 11.2\% in
2015, decreased to 8.5\% by 2017, with a 10/10 ratio of 31.6. However,
during the pandemic in 2020, poverty rose to 10.7\%, the 10/10 ratio
spiked to 301.7, and unemployment reached a decade-high of 10.5\%. By
2022, poverty declined to 6.5\%, the 10/10 ratio improved to 61.2, and
unemployment decreased to 7.9\%, with a slight rise to 8.7\% in 2023.

Chile implemented sanitary controls early in the pandemic, initiating
lockdowns on March 18, 2020, and lifting all restrictions by March 5,
2023. In this context, Chile stands out as one of the strictest
countries globally and within South America in terms of pandemic-related
policies. According to the Stringency Index (Hale et al. 2021)---ranging
from 0 to 100 (100 being the strictest) and encompassing measures like
school closures, workplace shutdowns, and travel bans---the global
median was 42.54, while Chile scored 58.1. This placed Chile 36th among
184 countries and 4th in South America, following Peru (60.1), Suriname
(60.5), and Venezuela (80.5). Besides, an important cycle of protests
known as the ``social outburst'' took place between October 2019 and
March 2020 (Somma et al. 2021). As a consequence there were two
constitutional processes happening in this very same period after the
massive social upheavals. However, the protest drastically decreased as
a result of the lockdown (Joignant et al. 2020).

In terms of public views on economic inequality, it has been documented
that critical views toward inequality increased in 2019 and deepened
under the pandemic (COES 2023), which we argue can be in part explained
by the increased risk exposure and vulnerability in the context of the
COVID-19 pandemic. Although the pandemic might have triggered such
concerns mostly regarding health, we believe that other policy areas,
such as education, could have been affected in this period.
Consistently, our first hypothesis is:

@@

Besides proposing a general decrease over time in the justification of
inequalities, a second hypothesis refers to specific aspects of the
pandemic. As some political measures from the state included several
social benefits (as delivering food boxes, free vaccination, direct
transfers, among others), we argue that being recipient of such benefits
could affect attitudes towards the redistributive role of the state as
equalizer opportunities to face risks. As a consequence, preferences for
educational inequalities would be reduced:

\(H_2\): Recipient of state benefits show lower justification of
educational inequalities

In the same line, we expect that those who give priority to take care of
the health of the population instead of the country economy would
exhibit larger preferences for educational inequalities:

\(H_3\): Those prioritizing health instead of the economic situation
show higher justification of educational inequalities

\emph{Meritocracy and educational inequalities}

Whether inequalities in education are considered just or not lead us to
take into account the concepts of merit and meritocracy. Meritocracy, a
concept coined by the sociologist Michael Young in his satirical work
``The Rise of the Meritocracy'' (1962), has a rich history and
multifaceted meaning. Young's term was initially used to criticize a
hypothetical future society where social and economic positions were
entirely determined by individuals' merits, talents, and efforts,
ostensibly eliminating the influence of inherited privilege. However,
over time, the term has evolved to describe a social system where
advancement and success are perceived as directly linked to one's
abilities and achievements, sidelining inequalities of opportunities
(Bell 2020). This would contribute to the justification of inequality in
the educational domain (Sandel 2020), as individual ability is often
considered inherited, and educational attainment is influenced by family
background (Goldin 2000).

Although there are few studies relating meritocratic beliefs and
inequality justification, from a broader perspective it is possible to
find research which incorporates the support of different distributive
principles with regard to education. For instance, in an experiment by
Igliozzi, Granot, and Ottati (2024), they establish how the support for
different justice principles (equity, equality, and need) is related to
support for redistributive policies in domains such as education. They
find that a larger system justification (Jost and Hunyady 2003) is
linked to higher support for the principles of equity and equality (over
need) in distributing educational outcomes, where equity was measured in
terms of access to education related to performance (`Students who place
within the highest 50\% of testing scores from the previous year can
attend the magnet school'). In the same line, Lee and Stacey (2023) show
that performance assessments (in what they call ``neoliberal
orientations'') predicted perceived fairness of educational inequality.

More specific empirical research examining the relationship between
meritocratic perceptions and the justification of educational inequality
provides valuable insights into the dynamics shaping individuals'
attitudes and beliefs. Some studies have explored the extent to which
people believe that educational systems are meritocratic and how these
perceptions influence their acceptance or rejection of educational
disparities, as well as status differences (McNamee and Miller 2004).
This belief is particularly influential in educational institutions,
where it not only encourages academic success but also legitimizes
social and income inequality (Batruch et al. 2023). Most of the evidence
so far suggests that those who believe in a direct link between effort,
talent, and educational success are more inclined to rationalize
existing disparities in educational outcomes. However, empirical
findings also reveal nuances in this relationship. For instance, studies
show that when individuals perceive structural barriers, such as
economic disparities or discrimination, as hindering the meritocratic
ideal, they may be less likely to justify educational inequalities (Day
and Norton 2023). This suggests that while meritocratic beliefs play a
significant role, individuals' awareness of broader societal issues can
mediate their willingness to rationalize educational disparities.

Empirical cross-sectional evidence from Chile reveals two significant
trends: low public support for the perception of meritocracy and limited
justification for unequal access to social services such as education
and health. For instance, Castillo, Madero-Cabib, and Salamovich (2013)
analyzed data from the International Social Survey Programme (ISSP) in
1999 and 2009, highlighting consistently low support for inequality
based on resource allocation or individual merit. Similarly, Castillo,
Miranda, and Carrasco (2012) demonstrated widespread skepticism among
Chileans regarding the meritocratic system, with many disagreeing with
the notion that rewards are fairly distributed based on talent and
effort. In line with this evidence, we propose that:

@@

Combining the arguments regarding the effect of the pandemic and its
relationship with meritocracy, we explore interactions between
meritocracy and inequality justification over time. As COVID-19 pandemic
has affected the structural opportunities to attend schools,
universities, or workplaces, the ideal of meritocracy might have been
challenged in the time frame analyzed. As a consequence, the
justification of inequality could show a decrease over time:

@@

Along with meritocratic perceptions, in this study we incorporate
inequality perception. As pointed out by Janmaat (2013), justifications
are usually confused under the labels of ``beliefs'' and ``attitudes,''
and their distinction from perceptions is relevant as it helps state
that justifying a certain level of inequality requires taking into
account which level of (perceived) inequality is first and foremost
being recognized (Son Hing et al. 2011; Castillo et al. 2023). For
instance, a recent study by Day and Norton (2023) on the perception and
justification of inequality in university endowments in the U.S. points
out that individuals tend to underestimate the magnitude of existing
inequality while simultaneously desiring greater equality. Furthermore,
through an experiment, they demonstrate that information about
inequality in university endowments (i.e., manipulating inequality
perception) increases the perception of injustice in education. Results
of this study are in line with the consistent evidence on biases related
to the underestimation of economic inequality (Castillo, García-Castro,
and Venegas 2022; Gimpelson and Treisman 2018). As a larger inequality
in this domain threatens a basic equality ideal, the corresponding
hypothesis in this case is:

\(H_6\): A larger inequality perception is negatively associated with
the justification of educational inequalities.

Next, we take into account the evidence that relates rational interests
to justifications and preferences (Son Hing et al. 2019). Research about
the justification of educational inequality usually focuses on
differences between status groups, based on the rational premise that
those who are worst-off would tend to show lower levels of justification
(Busemeyer and Iversen 2020). For instance, Valant and Newark (2016)
find that Americans are more concerned about --- and more supportive of
proposals to close --- wealth-based achievement gaps than Black-White or
Hispanic-White gaps. In line with this evidence, we expect that:

\(H_7\): Those with lower status characteristics show lower
justification of educational inequalities

@@

The main data source is the Chilean Longitudinal Social Survey ELSOC
(2016--2023). ELSOC has been designed as a yearly panel study to
evaluate how individuals think, feel, and behave regarding a set of
social issues concerning conflict and social cohesion in Chile. The
sampling design is probabilistic, stratified, clustered, and multistage.
It provides adequate coverage of the country's largest cities
(Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller
cities. The sample covers participants aged between 18 and 75 years (in
wave 1), representing 77\% of the total national population and 93\% of
the urban areas (ELSOC 2022).

The survey has been conducted yearly since 2016, except for the year
2020, when it was suspended due to the pandemic. The waves 2016, 2017,
2018, 2019, 2022, and 2023 were administered using computer-assisted
personal interviewing (CAPI). However, a reduced version was conducted
using computer-assisted telephone interviews (CATI) in 2021. In
addition, wave 3 included a refreshment sample to counter survey
attrition. As a result, the total sample of wave 3 included 3,748 cases,
of which 2,229 were part of the original sample and 1,519 were from the
refreshment sample. The data from the refreshment sample is not included
in the analytical sample for this study as we analyze longer response
trends. Regarding the original sample, the response rate was 62.4\% in
wave 1, achieving N = 2,927 participants. In broader terms, the
accumulated attrition between wave 1 and wave 7 is 40\%, achieving a
final sample of N = 1,741. The analysis used longitudinal weights to
avoid bias due to the sampling design, allowing us to control for biases
arising from systematic patterns of non-participation in the survey
after the first wave. For a more detailed analysis of responses,
attrition, and the construction of longitudinal weights, visit
https://coes.cl/encuesta-panel/. The dataset is publicly available at:
https://dataverse.harvard.edu/dataverse/elsoc.

@@

A third group of variables is included in the analysis to assess the
effect of elements related to the pandemic context. These variables were
only included in the 2021 wave, so each individual's responses were
replicated for the remaining years. The first variable addresses the
belief regarding whether, in the context of the pandemic, it is more
important to give priority to economic activity than to the health of
the population, measured on a Likert scale from (1) ``strongly
disagree'' to (2) ``strongly agree''. The second variable is a dummy
(No/Yes) that captures whether individuals received state benefits
during the pandemic (e.g., Food Box, Emergency Family Income,
Contribution for the Middle Class, or others).
